Limited plasticity but increased variance in physiological rates across ectotherm populations under climate change

### 2024-07-16 ms.qmd
Intro paragraph on why variance written, fixed a few things throughout, finished off discussion, mostly, just a few senetences need in one paragraph

Climate change will result in warmer and more variable temperatures globally impacting physiological rates and function in ectothermic animals. Acclimation of physiological rates can help maintain function, however, how variance in physiological rates change with temperature has been ignored despite its potential ecological and evolutionary importance. We apply new effect sizes that quantify the extent to which both mean and variance in physiological rates change as temperature increases. Meta-analysing >1900 effects from 226 species, we show that variance in physiological rates increases at higher temperatures, but that the magnitude of change depends on habitat. Freshwater and marine ectotherms are capable of acclimation and have the greatest increase in physiological rate variance. In contrast, terrestrial ectotherms lack a capacity for acclimation and have a smaller increase in variance. Simulations suggest fundamental differences in among-individual variation in thermal breadth and optima of performance curves across habitats that contribute to these differences. Our results highlight the greater vulnerability of terrestrial ectotherms to climate change because limited increases in physiological rates will provide less raw material for evolutionary adaptation. Considering both acclimation capacity and variance in physiological rates side-by-side is therefore important for understanding how climate change will impact populations.

@@

Reversible phenotypic plasticity, such as physiological acclimation, is relatively rapid and can be fine-tuned to environmental conditions making it the first ‘line-of-defense’ against environmental change [@Dewitt1998; @Scheiner2020]. For example, physiological rates are known to speed up as temperature increases because of the thermodynamic effects on chemical reaction rates -- so called 'acute' temperature responses (@fig-concept). However, longer-lasting (days-weeks) temperature increases that move environmental conditions away from thermal optima of physiological rate functions (i.e., thermal performance curves) can be mitigated by acclimation, which adjust reaction rates [@Seebacher2015-wi; @Havird2020-mj]. Physiological acclimation is driven by endocrine and epigenetic processes that change the underlying physiology to allow organisms to maintain physiological performance around a fitness optimum despite changes in the environment [@seebacher2019histone; @little2013thyroid; @taff2016endocrine]. Acclimation therefore alters acute thermal sensitivity to offset the potentially negative effects of acute temperature changes (e.g., higher energetic demands). Acclimation, however, does not normally result in physiological rates returning to their original state [‘complete’ compensation - *sensu* @huey1999testing]. Rather, increased physiological rates are only partially compensated [@huey1999testing; @Havird2020-mj].

Acclimation is expected to evolve in populations when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Thermal variation and predictability are known to differ across habitats (terrestrial, marine and freshwater) [@steele2019comparison], and as such, we expect organisms within these habitats to vary in their capacity for plastic adjustments in physiology. Species occupying terrestrial habitats have been shown to exhibit weak acclimation abilities, and so, are likely to be particularly vulnerable to climate change given their greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper; @morley2019physiological]. In contrast, marine and freshwater habitats appear to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022]. However, the focus of research up to now has been primarily on mean physiological responses neglecting how variability in physiological processes might also be impacted by higher temperatures.

As mean physiological rates increase with temperature there is reason to believe that the intrapopulation variability will also be impacted. Positive mean-variance relationships are common across biology suggesting that, as physiological rates increase with temperature, so to should variation [i.e., Taylor's Law; @giometto2015sample]. As biochemical reaction rates increase with temperatures

Changes in physiological rate variability is expected to have important consequences on the flow of energy within and between populations, communities, and ecosystems [@hendry2016key; @seebacher2023can; @barneche2021warming; @Sanderson2023; @Bolnick2011]. Generally, more variable populations are predicted to be associated with broader niches, have increased growth rate, and decreased vulnerability to environmental change lowering extinction risk [i.e., "portfolio effects", @forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key; @hart2016variation; @Schindler2010]. In addition, if phenotypic and genetic variation are correlated and traits are linked to fitness, changes in variation may limit responses to selection impacting a populations capacity to evolve [@Pelletier2012]. Therefore, maintaining intrapopulation variability in physiological rates in a warmer world may be important for population resilience to climate change.

Here, we use meta-analysis to re-evaluate the degree to which aquatic and terrestrial ectotherms are capable of physiological plasticity. We then developed new effect sizes to quantify how variance in physiological rates change with temperature to ask the following questions: 1) How much is variance in physiological rates expected to change, if at all, as temperatures rise? 2) Are temperature effects on plastic adjustments in physiological rates larger than changes in variance across aquatic and terrestrial ectotherms? 3) How does acclimation and changes in variance relate to different life-stages and types of traits, and does this depend on habitat? 4) Are changes in plasticity or variance in physiological rates impacted by past climate history? 5) How are means and variances in physiological rates expected to change under climate change?

@@

We extracted means, standard deviations, and sample sizes for physiological rates at the two test temperatures. If there were more than two test temperatures, we chose only the test temperatures that fell within the most likely natural range of temperatures experienced by the species in question. We extracted these data from text, tables or figures of a given paper. Data were extracted from figures using the R package *metaDigitise* [@Pick2019-mg]. We also recorded the phylum, class, order, genus and species, and the latitude and longitude from where the experimental animals were sourced. For studies that did not provide latitude and longitude for the population, we searched for similar studies by the same lab group to identify where the population was likely to have been sourced. If the experimental animals were derived from the wild, we recorded the nearest latitude and longitude of the field collection site. If the animals were sourced from a commercial supplier, we took the latitude and longitude of the supplier. When it was not possible to find latitude and longitude using these methods, we looked up the distribution of the species in question and took the latitude and longitude of the centroid of the species' distributional range.

@@

We recorded or derived a series of moderator variables from each study that are expected to have an impact on our effect size estimates. This included the duration of acclimation in days given that acclimation responses are expected to depend on how long chronic temperature exposure occurs (longer exposure = better acclimation response) [@Seebacher2015-wi]. We also recorded if the sample of animals were derived from captive or wild stocks, the life-history stage of the animals used ("adult" or "juvenile") and the habitat type ("freshwater", "marine" or "terrestrial") given that @Seebacher2015-wi show that these factors can impact $Q_{10}$ estimates. Physiological rate measures varied widely across the studies but could generally be grouped into two broad categories that included whole-organism measures, which all integrate a diversity of physiological and biochemical processes, and biochemical processes (e.g., enzyme reaction rates, proton leak) [@Seebacher2015-wi]. We explore differences across more detailed trait categories in *Supplemental Materials*, but note sample sizes are limited for many traits. Traits that could not be categorised into these two we classified as ‘Other’.

@@

$$
lnCVR_{Q_{1}} = \frac{lnCVR_{Q_{10}}}{10}
$$ {#eq-lCVRpred}

@@

Variance in physiological rates ($lnVR_{Q_{10}}$) showed an increase with increasing temperature across all habitat types ([@fig-2]B). Overall, there was a `r mean(t_VR_mean_ef)`% (95% CI: `r quantile(t_VR_mean_ef, 0.025)[1]` to `r quantile(t_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(t_VR_mean))`) increase in physiological rate variance for terrestrial ectotherms,  a `r mean(m_VR_mean_ef)`% (95% CI:`r quantile(m_VR_mean_ef, 0.025)[1]` to `r quantile(m_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(m_VR_mean))`) increase in variation for marine ectotherms and a `r mean(f_VR_mean_ef)`%, 95% CI: `r quantile(f_VR_mean_ef, 0.025)[1]` to `r quantile(f_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(f_VR_mean))` increase in variance for freshwater ectotherms across 10°C (@fig-2; results from "Model 2"). The relative differences between habitats were similar when considering $lnCVR_{Q_{10}}$ (See *Supplementary Materials*).

@@

Across habitats biochemical processes tended to result in greater increases in variance at higher temperatures, but this was only significant for marine ectotherms (biochemical/whole-organism contrasts: Marine: $\beta$ = `r contrast_lnvr_trait_cat_stats[2,3]`, 95% CI: `r contrast_lnvr_trait_cat_stats[2,4]` to `r contrast_lnvr_trait_cat_stats[2,5]`, $p_{MCMC}$ = `r contrast_lnvr_trait_cat_stats[2,6]`; Freshwater: $\beta$ = `r contrast_lnvr_trait_cat_stats[3,3]`, 95% CI: `r contrast_lnvr_trait_cat_stats[3,4]` to `r contrast_lnvr_trait_cat_stats[3,5]`, $p_{MCMC}$ = `r contrast_lnvr_trait_cat_stats[3,6]`; Terrestrial: $\beta$ = `r contrast_lnvr_trait_cat_stats[1,3]`, 95% CI: `r contrast_lnvr_trait_cat_stats[1,4]` to `r contrast_lnvr_trait_cat_stats[1,5]`, $p_{MCMC}$ = `r contrast_lnvr_trait_cat_stats[1,6]`) ([@fig-btrait]D-F; "Model 3"). Interestingly, variance increases for biochemical traits is reduced during acclimation in marine ectotherms ([@fig-btrait]E).

@@

We show that the capacity for acclimation of physiological rates differs across habitats. Our findings confirm previous results that quantify the different capacity of terrestrial, marine and freshwater ectotherms to acclimate [@Seebacher2015-wi; @Gunderson2015-hz; @morley2019physiological]. Our analysis generally supports that of @Seebacher2015-wi, @Gunderson2015-hz and @morley2019physiological that all show a general inability of terrestrial ectotherms to physiologically acclimate. These consistent results are surprising given the different physiological traits measured (i.e., thermal limits versus physiological rates). The change in acclimation $Q_{10}$ we found with our expanded dataset (34% increase in effects) was similar to @Seebacher2015-wi for freshwater organisms (~17%), but higher in marine ectotherms [decrease of 16% versus ~10% in @Seebacher2015-wi], and opposite in terrestrial ectotherms [increase of ~6% compared to an ~8% decrease in @Seebacher2015-wi], possibly due to the use of newly derived $Q_{10}$ effect sizes that allow us to control for sampling variance. Greater capacity for acclimation in aquatic organisms may be the result of fewer opportunities for behavioural thermoregulation in aquatic environments making physiological remodeling important for maintaining homeostasis [@Gunderson2015-hz; @morley2019physiological]. Importantly, even though marine and freshwater ectotherms were capable of partial acclimation, on average, the effect size was small (amounting to $Q_{10}$ dropping from ~1.8 to 1.6), suggesting acclimation will provide limited scope for aquatic ectotherms to adjust their physiology to higher temperatures.

Biochemical traits (e.g., enzyme activities) were more likely to acclimate compared to whole-organisms traits as predicted, but only for marine ectotherms. In contrast, both whole-organism and biochemical traits were equally likely to acclimate in freshwater ectotherms. Unsurprisingly, we observed no distinction between traits in terrestrial ectotherms......FRANK?

@@

Contrary to acclimation capacity variance in physiological rates increased across habitats with effect sizes being 3-5x larger than those observed for acclimation. Mechanistically, it is unclear what is contributing to the increased variation in physiological rates at higher temperatures, but it could be the result of increased among-individual variability in how biochemical, cellular and physiological processes function at higher temperatures to maintain homeostasis [@tattersall2012coping; @angilletta2009thermal; @somero1995proteins; @fields2001protein]. Higher temperatures increase membrane fluidity affecting electrochemical gradients and impacting protein structure and function [@somero1995proteins; @fields2001protein; @tattersall2012coping]. Such challenges (among others) may expose among-individual variation within a population. Indeed, even acclimation itself can impact among-individual variation [REF FRANK?], and we find evidence to that effect given that biochemical processes in marine organisms were both more likely to acclimate and have reduced variance increases compared to acute responses. However, this was not a general phenomena given that freshwater ectotherms were able to maintain high variance despite being able to acclimate.

Importantly, increased variance in physiological rates was not equal among ectotherms occupying terrestrial, marine and freshwater habitats, with increases in variance being higher in freshwater ectotherms (~60% increase / 10°C) compared to terrestrial ectotherms (~36% increase / 10°C). One possible hypothesis for the differences in variability we observed across habitats could be that among-individual variation in key parameters affecting the shape of thermal performance curves differ [@angilletta2009thermal; @huey1989evolution]. Our simulations suggest that theoretical and observed $lnVR_{Q_{10}}$ match when thermal performance curves have different among-individual variance in thermal optima and breadth making this hypothesis plausible. Such patterns across habitats are expected given that terrestrial ectotherms should be adapted to warmer and more variable environments. Theoretical models also suggest that populations with greater temporal environmental variability exhibit greater thermal breadth [@lynch1987environmental]. While we did not find support that thermal variation co-varied with $lnVR_{Q_{10}}$ (see below), as would be expected, such analyses will depend on an appropriate measure of temporal variation in temperature that is biologically relevant -- a challenging feat across diverse taxa but worthy of future investigation.

Our results further highlight the potential vulnerability of terrestrial ectotherms to climate change. Assuming changes in variation in physiological rates are underpinned by genetic variation, and there is a genetic correlation with fitness, smaller increases in physiological variance could limit adaptation in terrestrial habitats more than aquatic habitats in the future [@urban2023and]. For example, under climate change we expect an increase in variance in physiological rates of only ~13% in terrestrial habitats whereas for freshwater habitats we expect variation in physiological rates to increase by ~30%. There will obviously be limits to variance increases, and we predict that organisms closer to their upper thermal limits ($CT_{max}$) will have lower $lnVR_{Q_{10}}$ values compared to those farther away from $CT_{max}$. Some evidence points to possible differences across habitats in upper thermal limits already [@Gunderson2015-hz; @Pinsky2019-fn] making this a fruitful future question to ask.

@@

Acclimation capacities are expected to differ between life-stages because of distinct of patterns of dispersal, habitat use and behaviour that force earlier life stages to cope with more variable environmental conditions which can also lead to developmental constraints on how physiological systems respond later in life [@martin2015age; @stearns1976life; @angilletta2009thermal; @noble2018developmental; @o2019developmental; @Pottier2022; @sinclair2016can]. In addition, plastic responses are also expected to be costly [@angilletta2009thermal; @Dewitt1998], and such costs can be magnified in late life reducing the capacity for plasticity [e.g., @rossi2019ageing]. These processes can also result in changes to intrapopulation variation in physiological rates at higher temperatures but the direction of change between early and adult life stages is likely to depend on the costs of adjusting physiological processes, energy reserves at different life stages, and the extent to which early life experiences constrain plasticity.

Despite these expectations, more broadly, we show little difference between early and late life acclimation capacities and little change in the variance in physiological rates across habitats. This may not be too surprising given that such responses are likely context or trait dependent [@carter2020life; @moghadam2019heat]. The lack of differences we observed may be because both juvenile and adult animals occupy similar thermal niches, disperse to a similar extent and exhibit comparable thermoregulatory behaviors making physiological responses to temperature similar. It is possible that temperatures experienced earlier in development (e.g., egg or larval stages) exhibit greater capacity for acclimation because of limited dispersal capabilities and very different habitats. Some empirical evidence supports this in insects. For example, @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults. A focus on collecting more detailed information on behaviour, dispersal and thermal environments experienced by different life stages is likely to provide a more complete picture on when plasticity differs. To date, however, there are few studies that have focused on how variability in responses change. We would also encourage more empirical focus on this questions given the potential ecological and evolutionary implications.

@@

Enhanced knowledge of how variation in physiological rates vary across populations and species, and the degree to which they can be adjusted in response to the environment leads to more informed predictions about the ecological and evolutionary dynamics of natural populations [@forsman2015rethinking; @seebacher2023can; @cooke2021one; @Sanderson2023]. We show general patterns across taxa and habitats that provide a foundation to understand the relationship between plasticity and trait variance, as well as particular trade-offs that could impact the benefits (or lack thereof) of acclimation. It is important to recognise, however, that these patterns do not necessarily apply to all populations. Substantial variation in acclimation responses and changes in variance exist among populations and traits, as evidenced by wide prediction intervals and substantial study- and trait-level variance estimates. Conservation efforts are often targeted at particular populations or species, and taxonomic differences are important in this context. Regardless, quantitative measures of the changes in variance in physiological rates could be better incorporated into physiological and ecological models to provide more nuanced, and possibly more realistic, predictions about the impacts of climate change on natural populations. While we do not yet understand the relative contribution of environmental and genetic factors to variance changes, models could better decouple how different levels of heritability and total variance impact evolutionary and ecological predictions. Our meta-analysis now provides the opportunity to parameterise such models, and ensure they are better aligned with empirical findings.

@@

We would like to thank Malgorzata Lagisz for helping sort out the phylogeny used in our analysis and Enrico Rezende for discussions around the implementation of thermal performance curves. We are also grateful to three anonymous reviewers who all provided highly constructive feedback that greatly improved the manuscript.

@@

figs3 <- (ps3_1 | ps3_2 | ps3_3) + plot_annotation(tag_levels = "A", tag_suffix = ")") & theme(plot.tag = element_text(size = 20))
ggsave(figs3, width = 16, height = 5.7, filename = "./output/figures/figS12.png")

figs3 <- image_read("./output/figures/figS12.png")
figs3
```

>>>>

Climate change will result in warmer and more variable temperatures globally impacting physiological rates and function in ectothermic animals. Acclimation of physiological rates can help maintain function, however, how variance in physiological rates change with temperature has been ignored despite its potential ecological and evolutionary importance. We apply new effect sizes that quantify the extent to which both mean and variance in physiological rates change as temperature increases. Meta-analysing >1900 effects from 226 species, we show that variance in physiological rates increases at higher temperatures, but that the magnitude of change depends on habitat. Freshwater and marine ectotherms are capable of acclimation and have the greatest increase in physiological rate variance. In contrast, terrestrial ectotherms lack a capacity for acclimation and have a smaller increase in variance. Simulations suggest fundamental differences in among-individual variation in thermal breadth and optima of performance curves across habitats that may contribute to these differences. Our results highlight the greater vulnerability of terrestrial ectotherms to climate change because limited increases in physiological rates may provide less raw material for evolutionary adaptation. Considering both acclimation capacity and variance in physiological rates side-by-side is therefore important for understanding how climate change will impact populations.

@@

Reversible phenotypic plasticity, such as physiological acclimation, is relatively rapid and can be fine-tuned to environmental conditions making it the first ‘line-of-defense’ against environmental change [@Dewitt1998; @Scheiner2020]. For example, physiological rates are known to speed up as temperature increases because of the thermodynamic effects on chemical reaction rates -- so called 'acute' temperature responses (@fig-concept). However, longer-lasting (days-weeks) temperature increases that move environmental conditions away from thermal optima of physiological rate functions (i.e., thermal performance curves) can be mitigated by acclimation, which adjust reaction rates [@Seebacher2015-wi; @Havird2020-mj]. Physiological acclimation is driven by endocrine and epigenetic processes that change the underlying physiology to allow organisms to maintain physiological performance around a fitness optimum despite changes in the environment [@seebacher2019histone; @little2013thyroid; @taff2016endocrine]. Acclimation therefore alters acute thermal sensitivity to offset the potentially negative effects of acute temperature changes (e.g., higher energetic demands). Acclimation, however, does not normally result in physiological rates returning to their original state [‘complete’ compensation - *sensu* @huey1999testing]. Rather, increased physiological rates are usually only partially compensated [@huey1999testing; @Havird2020-mj].

Acclimation is expected to evolve in populations when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Thermal variation and predictability are known to differ across habitats (terrestrial, marine and freshwater) [@steele2019comparison], and as such, we expect organisms within these habitats to vary in their capacity for plastic adjustments in physiology. @rohr2018complex show relationships between acclimation capacity, latitude and body size suggesting climate could be an important driver of acclimation responses. In addition, species occupying terrestrial habitats have been shown to exhibit weaker acclimation abilities, and so, may be particularly vulnerable to climate change given their greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper; @morley2019physiological]. In contrast, marine and freshwater habitats appear to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022]. However, the focus of research up to now has been primarily on mean physiological responses neglecting how variability in physiological processes might also be impacted by higher temperatures.

As mean physiological rates increase with temperature there is reason to believe that intrapopulation variability will also be impacted. Positive mean-variance relationships are common across biology suggesting that, as physiological rates increase with temperature, so to should variation [i.e., Taylor's Law; @giometto2015sample]. Differences in the shape of thermal performance curves (thermal breadth, maximal performance and thermal optima) can exacerbate among-individual variability at higher temperatures, which can also depend on different levels of biological organisation, environmental conditions and acclimation responses [@rezende2019thermal; @tattersall2012coping; @schulte2011thermal; @angilletta2009thermal]. Presumably, increases in underlying variation in physiological rates should be related to varying robustness among individuals in how biochemical and physiological networks are affected by temperature [@schulte2011thermal]. A focus on quantifying the shape of performance curves has often taken precedence over quantifying levels of among-individual variation in curves which has limited our understanding of their capacity to evolve.

Importantly, changes in physiological rate variability is expected to have consequences on the flow of energy within and between populations, communities, and ecosystems [@hendry2016key; @seebacher2023can; @barneche2021warming; @Sanderson2023; @Bolnick2011]. Generally, more variable populations are predicted to be associated with broader niches, have increased growth rate, and decreased vulnerability to environmental change lowering extinction risk [i.e., "portfolio effects",@Schindler2010] [see also, @forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key; @hart2016variation; @Schindler2010; @portner2021climate]. In addition, if phenotypic and genetic variation in physiological rates are correlated rates and linked to fitness, changes in variation may limit responses to selection impacting a populations capacity to evolve [@Pelletier2012; @hoffmann2011climate]. Therefore, maintaining intrapopulation variability in physiological rates in a warmer world may be important for population resilience to climate change.

Here, we use meta-analysis to re-evaluate the degree to which aquatic and terrestrial ectotherms are capable of physiological plasticity. We then developed new effect sizes to quantify how variance in physiological rates change with temperature to ask the following questions: 1) How much is variance in physiological rates expected to change, if at all, as temperatures rise? 2) Are temperature effects on plastic adjustments in physiological rates larger than changes in variance across aquatic and terrestrial ectotherms? 3) How does acclimation and changes in variance relate to different life-stages and types of traits, and does this depend on habitat? 4) Are changes in plasticity or variance in physiological rates impacted by past climate history? 5) How are variances in physiological rates expected to change under climate change?

@@

We extracted means, standard deviations, and sample sizes for physiological rates at the two test temperatures ([@fig-concept]A). If there were more than two test temperatures, we chose only the test temperatures that fell within the most likely natural range of temperatures experienced by the species in question (@fig-concept). We extracted these data from text, tables or figures of a given paper. Data were extracted from figures using the R package *metaDigitise* [@Pick2019-mg]. We also recorded the phylum, class, order, genus and species, and the latitude and longitude from where the experimental animals were sourced. For studies that did not provide latitude and longitude for the population, we searched for similar studies by the same lab group to identify where the population was likely to have been sourced. If the experimental animals were derived from the wild, we recorded the nearest latitude and longitude of the field collection site. If the animals were sourced from a commercial supplier, we took the latitude and longitude of the supplier. When it was not possible to find latitude and longitude using these methods, we looked up the distribution of the species in question and took the latitude and longitude of the centroid of the species' distributional range.

@@

We recorded or derived a series of moderator variables from each study that are expected to have an impact on our effect size estimates. This included the duration of acclimation in days given that acclimation responses are expected to depend on how long chronic temperature exposure occurs (longer exposure = better acclimation response) [@Seebacher2015-wi]. We also recorded if the sample of animals were derived from captive or wild stocks, the life-history stage of the animals used ("adult" or "juvenile") and the habitat type ("freshwater", "marine" or "terrestrial") given that @Seebacher2015-wi show that these factors can impact $Q_{10}$ estimates. Physiological rate measures varied widely across the studies but could generally be grouped into two broad categories that included whole-organism measures, which all integrate a diversity of physiological and biochemical processes, and biochemical processes (e.g., enzyme reaction rates, proton leak) [@Seebacher2015-wi; @rezende2019thermal]. We explore differences across more detailed trait categories in *Supplemental Materials*, but note sample sizes are limited for many traits. Traits that could not be categorised into these two we classified as ‘Other’.

@@

$$
lnVR_{Q_{1}} = \frac{lnVR_{Q_{10}}}{10}
$$ {#eq-lCVRpred}

@@

Variance in physiological rates ($lnVR_{Q_{10}}$) showed an increase with increasing temperature across all habitat types ([@fig-2]B). Overall, there was a `r mean(t_VR_mean_ef)`% (95% CI: `r quantile(t_VR_mean_ef, 0.025)[1]` to `r quantile(t_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(t_VR_mean))`) increase in physiological rate variance for terrestrial ectotherms,  a `r mean(m_VR_mean_ef)`% (95% CI:`r quantile(m_VR_mean_ef, 0.025)[1]` to `r quantile(m_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(m_VR_mean))`) increase in variation for marine ectotherms and a `r mean(f_VR_mean_ef)`% (95% CI: `r quantile(f_VR_mean_ef, 0.025)[1]` to `r quantile(f_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(f_VR_mean))`) increase in variance for freshwater ectotherms across 10°C (@fig-2; results from "Model 2"). The relative differences between habitats were even larger when considering $lnCVR_{Q_{10}}$, which controls for changes in mean rates (See *Supplementary Materials*).

@@

Across habitats biochemical processes tended to result in greater increases in variance at higher temperatures, but this was only significant for marine ectotherms (biochemical/whole-organism contrasts: Marine: $\beta$ = `r contrast_lnvr_trait_cat_stats[2,3]`, 95% CI: `r contrast_lnvr_trait_cat_stats[2,4]` to `r contrast_lnvr_trait_cat_stats[2,5]`, $p_{MCMC}$ = `r contrast_lnvr_trait_cat_stats[2,6]`; Freshwater: $\beta$ = `r contrast_lnvr_trait_cat_stats[3,3]`, 95% CI: `r contrast_lnvr_trait_cat_stats[3,4]` to `r contrast_lnvr_trait_cat_stats[3,5]`, $p_{MCMC}$ = `r contrast_lnvr_trait_cat_stats[3,6]`; Terrestrial: $\beta$ = `r contrast_lnvr_trait_cat_stats[1,3]`, 95% CI: `r contrast_lnvr_trait_cat_stats[1,4]` to `r contrast_lnvr_trait_cat_stats[1,5]`, $p_{MCMC}$ = `r contrast_lnvr_trait_cat_stats[1,6]`) ([@fig-btrait]D-F; "Model 3"). Variance increases for biochemical traits is reduced during acclimation in marine ectotherms ([@fig-btrait]E).

@@

We show that the capacity for acclimation of physiological rates differs across habitats. Our findings confirm previous results that quantify the different capacity of terrestrial, marine and freshwater ectotherms to acclimate [@Seebacher2015-wi; @Gunderson2015-hz; @morley2019physiological]. Our analysis generally supports that of @Seebacher2015-wi, @Gunderson2015-hz and @morley2019physiological that all show a general inability of terrestrial ectotherms to physiologically acclimate. These consistent results are interesting given the different physiological traits measured in these meta-analyses (i.e., thermal limits versus physiological rates).

The change in acclimation $Q_{10}$ we found with our expanded dataset (34% increase in effects) was similar to @Seebacher2015-wi for freshwater organisms (~17%), but higher in marine ectotherms [decrease of 16% versus ~10% in @Seebacher2015-wi], and opposite in terrestrial ectotherms [increase of ~6% compared to an ~8% decrease in @Seebacher2015-wi], possibly due to the use of newly derived $Q_{10}$ effect sizes that allow us to control for sampling variance. Greater capacity for acclimation in aquatic organisms may be the result of fewer opportunities for behavioural thermoregulation in aquatic environments making physiological remodeling important for maintaining homeostasis [@Gunderson2015-hz; @morley2019physiological]. Importantly, even though marine and freshwater ectotherms were capable of partial acclimation, on average, the effect size was small (amounting to $Q_{10}$ dropping from ~1.8 to 1.6), suggesting acclimation will provide limited scope for aquatic ectotherms to adjust their physiology to higher temperatures.

Biochemical traits (e.g., enzyme activities) were more likely to acclimate compared to whole-organisms traits as predicted, but only for marine ectotherms, presumably because of the less complicated physiological networks that underlie specific biochemical processes [@schulte2011thermal]. However, this prediction did not hold up generally with both whole-organism and biochemical traits being equally likely to acclimate in freshwater ectotherms, which may be due to the different types of biochemical processes measured in the two habitats.

@@

Contrary to acclimation capacity variance in physiological rates increased across habitats with effect sizes being 3-5x larger than those observed for acclimation. Mechanistically, it is unclear what exactly is contributing to the increased variation in physiological rates at higher temperatures, but it is likely the result of increased among-individual variability in how biochemical, cellular and physiological processes function at higher temperatures to maintain homeostasis [@tattersall2012coping; @angilletta2009thermal; @somero1995proteins; @fields2001protein; @schulte2011thermal]. Higher temperatures increase membrane fluidity affecting electrochemical gradients and impacting protein structure and function [@somero1995proteins; @fields2001protein; @tattersall2012coping]. Such challenges (among others) may expose among-individual variation within a population. Indeed, even acclimation itself can impact among-individual variation, possibly through changes in individual performance curves [@schulte2011thermal]. Indeed, we find evidence to that effect given that biochemical processes in marine organisms were both more likely to acclimate and have reduced variance increases compared to acute responses. However, this was not a general phenomena given that freshwater ectotherms were able to maintain high variance despite being able to acclimate.

Importantly, increased variance in physiological rates was not equal among terrestrial, marine and freshwater ectotherms, with increases in variance being higher in freshwater ectotherms (~60% increase / 10°C) compared to terrestrial ectotherms (~36% increase / 10°C). One possible hypothesis for the differences in variability we observed across habitats could be that among-individual variation in key parameters affecting the shape of thermal performance curves differ [@angilletta2009thermal; @huey1989evolution; @tattersall2012coping; @rezende2019thermal]. Our simulations suggest that theoretical and observed $lnVR_{Q_{10}}$ match when thermal performance curves have different among-individual variance in thermal optima and breadth across habitats making this hypothesis plausible. Such patterns across habitats are expected given that terrestrial ectotherms should be adapted to warmer and more variable environments. Theoretical models also suggest that populations with greater temporal environmental variability exhibit greater thermal breadth [@lynch1987environmental]. While we did not find support that thermal variation co-varied with $lnVR_{Q_{10}}$ (see below), as would be expected, such analyses will depend on an appropriate measure of temporal variation in temperature that is biologically relevant -- a challenging feat across diverse taxa, but worthy of future investigation.

Our results further highlight the potential vulnerability of terrestrial ectotherms to climate change. Assuming changes in variation in physiological rates are underpinned by genetic variation, and there is a genetic correlation with fitness, smaller increases in physiological variance could limit adaptation in terrestrial habitats more than aquatic habitats in the future [@urban2023and; @hoffmann2011climate]. For example, under climate change we expect an increase in variance in physiological rates of only ~13% in terrestrial habitats whereas for freshwater habitats we expect variation in physiological rates to increase by ~30%. There will obviously be limits to variance increases, and we predict that organisms closer to their upper thermal limits ($CT_{max}$) will have lower $lnVR_{Q_{10}}$ values compared to those farther away from $CT_{max}$. Some evidence points to possible differences across habitats in upper thermal limits already [@Gunderson2015-hz; @Pinsky2019-fn] making this a fruitful future question to explore.

@@

Acclimation capacities are expected to differ between life-stages because of distinct patterns of dispersal, habitat use and behaviour that force earlier life stages to cope with more variable environmental conditions which can also lead to developmental constraints on how physiological systems respond later in life [@martin2015age; @stearns1976life; @angilletta2009thermal; @noble2018developmental; @o2019developmental; @Pottier2022; @sinclair2016can]. In addition, plastic responses are also expected to be costly [@angilletta2009thermal; @Dewitt1998], and such costs can be magnified in late life reducing the capacity for plasticity [e.g., @rossi2019ageing]. These processes can also result in changes to intrapopulation variation in physiological rates at higher temperatures but the direction of change between early and adult life stages is likely to depend on the costs of adjusting physiological processes, energy reserves at different life stages, and the extent to which early life experiences constrain plasticity.

Despite these expectations, more broadly, we show little difference between early and late life acclimation capacities and little change in the variance in physiological rates across habitats. This may not be too surprising given that such responses are likely context or trait dependent [@carter2020life; @moghadam2019heat]. The lack of differences we observed may be because both juvenile and adult animals occupy similar thermal niches, disperse to a similar extent and exhibit comparable thermoregulatory behaviors making physiological responses to temperature similar. It is possible that temperatures experienced earlier in development (e.g., egg or larval stages) exhibit greater capacity for acclimation because of limited dispersal capabilities and very different habitats. Some empirical evidence supports this in insects. For example, @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults. A focus on collecting more detailed information on behaviour, dispersal and thermal environments experienced by different life stages is likely to provide a more complete picture on when plasticity differs. To date, however, there are few studies that have focused on how variability in responses change. We would also encourage more empirical focus on this question given its potential ecological and evolutionary implications.

@@

Enhanced knowledge of how variation in physiological rates vary across populations and species, and the degree to which they can be adjusted in response to the environment leads to more informed predictions about the ecological and evolutionary dynamics of natural populations [@forsman2015rethinking; @seebacher2023can; @cooke2021one; @Sanderson2023]. We show general patterns across taxa and habitats that provide a foundation to understand the relationship between plasticity and trait variance, as well as particular trade-offs that could impact the benefits (or lack thereof) of acclimation. It is important to recognise, however, that these patterns do not necessarily apply to all populations. Substantial variation in acclimation responses and changes in variance exist among populations and traits, as evidenced by wide prediction intervals and substantial study- and trait-level variance estimates, which is consistent with our understanding of factors influencing variation in performance curves across taxa [@rezende2019thermal; @tattersall2012coping]. Conservation efforts are often targeted at particular populations or species, and taxonomic differences are important in this context. Regardless, quantitative measures of the changes in variance in physiological rates could be better incorporated into physiological and ecological models to provide more nuanced, and possibly more realistic, predictions about the impacts of climate change on natural populations. While we do not yet understand the relative contribution of environmental and genetic factors to variance changes, models could better decouple how different levels of heritability and total variance impact evolutionary and ecological predictions. Our meta-analysis now provides the opportunity to parameterise such models, and ensure they are better aligned with empirical findings.

@@

We would like to thank Malgorzata Lagisz for helping sort out the phylogeny used in our analysis and Enrico Rezende for discussions around alternative implementations of thermal performance curves. We are also grateful to three anonymous reviewers who all provided extremely constructive feedback that greatly improved the manuscript.

@@

figS12 <- (ps3_1 | ps3_2 | ps3_3) + plot_annotation(tag_levels = "A", tag_suffix = ")") & theme(plot.tag = element_text(size = 20))
ggsave(figS12, width = 16, height = 5.7, filename = "./output/figures/figS12.png")

figS12 <- image_read("./output/figures/figS12.png")
figS12
```
